---
title: "General Social Survey"
author: "Joel Calderin"
date: "Fall 2023"
output: 
  html_document:
    number_sections: true
    toc: true
---


```{r include=FALSE,echo=FALSE}
require(tigerstats)
require(tidyverse)
require(foreign)
source("GrabData.R")
```




#  Question 1

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data in the dataframe GSSdata is from the 2018 General Social Survey. The first blocks of R-code has selected down a subset of the data to just 16 variables. It has further removed unwanted factor levels in much of the data. Examine the code in the GrabData.R file to see what it is doing. Some of the variables are categorical and others are numerical. Be sure to do a variable analysis before tackling each question.  
First question - Is opinion on the death penalty (CAPPUN) independent of gun ownership (OWNGUN)?

$H_0$ There is no relationship between gun ownership and the opinion on the death penalty.

$H_A$ There is a relationship between gun ownership and the opinion on the death penalty.

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

Both of the variables are categorical. Therefore, we will use the Cat~Cat helper file. 

## Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
barchartGC(~OWNGUN + CAPPUN,data=GSSdata)
barchartGC(~OWNGUN + CAPPUN,data=GSSdata, type="percent")
```


This graph shows that the gun owner are more likely to favor the death penalty. Roughly 80% of gun owners approved of the death penalty, while roughly 20% of gun owners opposed of the death penalty. 


#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~OWNGUN + CAPPUN, data=GSSdata)
rowPerc(table1)
colPerc(table1)
```


This summary shows that gun owners are more likely to be in favor of the death penalty. Among the gun owners, 75.68% were in favor, while 24.32% opposed the death penalty.


### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```


The chi-squared test shows a P-Value of 0.02022 which is less than 0.05. This means we reject the NULL Hypothesis. 

The fisher exact test gave us a P-Value of 0.01 which is also less than 0.05. Therefore we reject the NULL hypothesis. The test also tells us that people that own a gun are on average 2.3 times more likely to favor capital punishment (death penalty) than those who don't favor.  

#  Question 2

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Is there a difference in years of education (EDUC) between the those who did or did not vote in 2012 (VOTE12)?

$H_0$ There is no difference in years of education between those who did or did not vote in 2012. 

$H_A$ There is a difference in the years of education between those who did or did not vote in 2012.

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

The years of education is a numerical variable, and votes are a categorical variable. Therefore I will be using the Numerical~Cat helper file.

## Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
plot(EDUC~VOTE12,data=GSSdata)
```

The graph shows that people with a higher education voted more in 2012 compared to those who did not vote. The median of those who voted is higher than the median of those who did not vote.



#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
favstats(EDUC~VOTE12,data=GSSdata)
```

The data shows that the average of people that voted in 2012 is  higher than the average of people that did not vote in 2012. 



### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->


```{r}
model1 <- aov(EDUC~VOTE12,data=GSSdata)
summary.aov(model1)
```

The P-Value is 0.000000453 which is significantly lower than 0.05. Therefore we reject the NULL hypothesis.

#  Question 3

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Determine if a persons performance on the wordsum test (WORDSUM) is independent of their level of schooling (DEGREE).

$H_0$ There is no difference in the WORDSUM test score based on level of schooling.

$H_A$ There is a difference in the WORDSUM test score based on level of schooling. 

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

The WORDSUM test scores are numerical data, and the level of schooling (DEGREE) is a categorical variable. Therefore we use the Numerical~Cat helper file. 

## Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
plot(WORDSUM~DEGREE,data=GSSdata)
```

The graph shows that as the degree level gets higher, the score on the WORDSUM test was higher. The highest score was Graduate Degree, while the lowest score was the LT high school degree. 

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
favstats(WORDSUM~DEGREE,data=GSSdata)
```

The data shows that the highest average score for the WORDSUM test was for the Graduate degree, while the lowest average score was for the LT High School Degree. Junior College and Bachelor degree averaged around the same score. 

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
model1 <- aov(WORDSUM~DEGREE,data=GSSdata)
summary.aov(model1)
```

The P-Value is 0.00000025 which is significantly less than 0.05. Therefore we will reject the NULL hypothesis.

# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->

